Case study validations of automatic bullet matching\tnoteref{t1}\tnoteref{t2}

### 2019-11-10 index.Rnw
Abstract changes

\tnotetext[t2]{The authors wish to thank Alan Zheng (NIST), Tylor Klep (Phoenix PD), and Melissa Nally and Kasi Kirksey (FSI Houston) for access to test set bullets. We would also like to thank the efforts of the Roy J Carver High Resolution lab in scanning the bullet sets and providing the scans to us. Finally, thanks to Prof James Hamby for his advice on XXX.}

@@

\author[isu,csafe]{Heike Hofmann\corref{corauthor}}
\cortext[corauthor]{Corresponding author}\ead{hofmann@iastate.edu}
\author[isu,csafe]{Susan Vanderplas}
\address[isu]{Statistics Department, Iowa State University\\2438 Osborne Dr, Ames, IA 50011}
\address[csafe]{Center for Statistical Applications in Forensic Evidence, Iowa State University\\613 Morrill Rd, Ames, IA 50011}

\begin{abstract}
Recent advances in microscopy have made it possible to collect 3D topographic data, enabling virtual comparisons based on the collected 3D data next to traditional comparison microscopy. Automatic matching algorithms have been introduced for various scenarios, such as matching cartridge cases (Tai and Eddy 2018) or matching bullet striae (Hare et al. 2017b,  Chu et al 2013,  De Kinder and Bonfanti 1999). One key aspect of validating automatic matching algorithms is to evaluate the performance of the algorithm on external tests. Here, we are presenting a discussion of the performance of the matching algorithm (Hare et al. 2017b) in three studies. We are considering matching performance based on the Random forest score, cross correlation, and consecutive matching striae (CMS) at the land-to-land level and, using Sequential Average Maxima scores, also at the bullet-to bullet level. Cross correlation and Random Forest scores both result in perfect discrimination of same-source and different-source bullets. At the land-to-land level, discrimination (based on area under the curve, AUC) is excellent ($\ge$ 0.90).
\end{abstract}

@@

In the matching algorithm proposed in \citet{aoas}, several features are combined using a random forest \citep{breiman} to produce a similarity score based on 3D topographic scans of land engraved areas (LEAs). In order to generate these features, some pre-processing is necessary in order to transform the 3D surface measurements into `signatures' which can be compared.

@@

From each land engraved area a signature is derived using the process described in detail in \citet{aoas}:

@@

Once signatures have been extracted from stable regions with expressed striae, they must be aligned in order to assess their match strength. Maximized cross correlation is used to pair-wise align signatures \citep{aoas,vorburgerApplicationsCrosscorrelationFunctions2011}.

@@

The random forest presented in \citet{aoas} is based on a combination of multiple characteristics, including cross correlation, number of matching striae, and number of consecutively matching striae. The output of the random forest is a score between 0 and 1 representing the algorithm's assessment of the strength of the match, where scores close to 1 indicate it to be highly likely that the aligned signatures come from the same source, while scores close to 0 indicate that the aligned signatures are highly likely to originate from different sources.

@@

When evaluating the \citet{aoas} algorithm, we will describe errors as a percentage of correct evaluations. Statistics that do not translate to percent-form interpretation, such as area under the curve (AUC), will be presented as a decimal.

@@

In this paper, we validate the algorithm described in \citet{aoas} on three external test sets. We assess the results on each set and evaluate whether the minimal requirements described above are fulfilled for bullet-to-bullet SAM scores. We compare the random forest algorithm's performance to the performance of other suggested measures for quantitative assessment of match strength, such as cross correlation and consecutive matching striae.

@@

The algorithm in \citet{aoas} is trained on scans from Hamby sets 252 and 173 made available through \citet{nist}. Set 173 was originally published as Hamby set 44. This mis-labeling has been corrected now. The scans used to train the model were taken at 20 fold magnification for a resolution of 1.5625 microns per pixel. While magnification is generally of interest in microscopy, **resolution** -- usually measured in microns per pixel -- is of more interest for 3D topographic measurements, as it determines the operative level of available data.

@@

Case studies were chosen such that difficulty for the matching algorithm increases with each case study: Hamby set 44 is part of the Hamby study. The algorithm in \citet{aoas} was trained on sets 173 and 252, so the bullets in Hamby 44 are of the same type of ammunition and are fired through the same barrels as the bullets in the training set. The Phoenix PD set uses Ruger P-95 barrels, which are different from the barrels in the Hamby sets used for training the algorithm, but the barrels are rifled similarly to the Ruger P-85 barrels. The Houston sets use Ruger LCP barrels. These barrels are first rifled traditionally, i.e. similar to the Ruger P-85, but are treated to a secondary round of heating after rifling. This second round of heat exposure has the potential to introduce some subclass characteristics, i.e.\ anomalies that are shared between different barrels manufactured in this manner. This should make the automatic matching harder, and in particular, is expected to complicate the classification as different-source land engraved areas/bullets. In all three studies, the scan resolution is much higher than the scans used to train the algorithm in \citet{aoas}; this difference also provides a test of the algorithm's ability to generalize to scans taken at different resolutions.

@@

At the beginning of this study, we anticipated Hamby 44 would be the easiest set to work with because of its similarity to the sets used to train the random forest algorithm. In a surprise turn of events, it was the hardest to solve, in part because of damage to the bullets that obscured LEA striae. It has been shown \citet{lpr} that parts of lands can be used for successful identifications. The algorithm proposed by \citet{aoas} is not capable of  automatically detecting parts of lands with well-expressed striae. It may be useful to couple the \citet{aoas} algorithm with an algorithm which assesses the quality of the input data and determines which portions of the data to use for comparison. This would emulate the process used by examiners, who first assess the quality of the evidence and whether there is enough information present to attempt a match. One major disadvantage of an automated algorithm is that all of the decisions humans make (recognizing degraded land areas, excluding those areas from consideration, matching only the remaining areas) must be explicitly characterized; however, this explicit characterization means that the process can be scientifically validated to a much higher degree than the human perceptual process.

@@

The random forest matching algorithm presented in \citet{aoas} does perform well on three different external test sets. While a single cutoff value cannot be used to distinguish matches and non-matches across the different test sets, the algorithm makes no errors when a set-specific cutoff value is used on bullet-to-bullet scores aggregated using sequence average maximum. This performance is on par with the performance of forensic examiners. For a future round of the Houston study, we are planning to compare the algorithmic performance directly with scores given by forensic toolmark examiners. \hh{XXX need to ask XXX  } The algorithm's score is not intended to replace an examiner's judgment; instead, it provides a complementary tool to measure and quantify the strength of an identification. 
By validating the algorithm on external test sets, we have demonstrated that the method can be generalized to different types of ammunition and is not overly sensitive to small differences in rifling procedure. Future studies can and should generalize this to a wider range of external test sets to establish the limits of the algorithm's generalizability.

>>>>

\tnotetext[t2]{The authors wish to thank Alan Zheng (NIST), Tylor Klep (Phoenix PD), and Melissa Nally and Kasi Kirksey (FSI Houston) for access to test set bullets. We would also like to thank the efforts of the Roy J Carver High Resolution lab in scanning the bullet sets and providing the scans to us.}

@@

\author[isu,csafe]{Heike Hofmann}
\cortext[corauthor]{Corresponding author}\ead{srvander@iastate.edu}
\author[isu,csafe]{Susan Vanderplas\corref{corauthor}}
\address[isu]{Statistics Department, Iowa State University\\2438 Osborne Dr, Ames, IA 50011}
\address[csafe]{Center for Statistical Applications in Forensic Evidence, Iowa State University\\613 Morrill Rd, Ames, IA 50011}

\begin{abstract}
Recent advances in microscopy have made it possible to collect 3D topographic data, enabling virtual comparisons based on the collected 3D data as a supplement to traditional comparison microscopy and 2D photography. Automatic matching algorithms have been introduced for various scenarios, such as matching cartridge cases \citep{tai2018fully,song3DTopographyMeasurements2014} or matching bullet striae \citep{aoas2, chuAutomaticIdentificationBullet2013, dekinderAutomatedComparisonsBullet1999}. One key aspect of validating automatic matching algorithms is to evaluate the performance of the algorithm on external tests, that is, using data which were not used to train the algorithm. Here, we present a discussion of the performance of the matching algorithm \citep{aoas2} in three studies conducted using different Ruger weapons. We consider the performance of three scoring measures: random forest score, cross correlation, and consecutive matching striae (CMS) at the land-to-land level and, using Sequential Average Maxima scores, also at the bullet-to bullet level. Cross correlation and random forest scores both result in perfect discrimination of same-source and different-source bullets. At the land-to-land level, discrimination for both cross correlation and random forest scores (based on area under the curve, AUC) is excellent ($\ge$ 0.90).
\end{abstract}

@@

In the matching algorithm proposed in \citet{aoas2}, several features are combined using a random forest \citep{breiman} to produce a similarity score based on 3D topographic scans of land engraved areas (LEAs). In order to generate these features, some pre-processing is necessary in order to transform the 3D surface measurements into `signatures' which can be compared.

@@

From each land engraved area a signature is derived using the process described in detail in \citet{aoas2}:

@@

Once signatures have been extracted from stable regions with expressed striae, they must be aligned in order to assess their match strength. Maximized cross correlation is used to pair-wise align signatures \citep{aoas2,vorburgerApplicationsCrosscorrelationFunctions2011}.

@@

The random forest presented in \citet{aoas2} is based on a combination of multiple characteristics, including cross correlation, number of matching striae, and number of consecutively matching striae. The output of the random forest is a score between 0 and 1 representing the algorithm's assessment of the strength of the match, where scores close to 1 indicate it to be highly likely that the aligned signatures come from the same source, while scores close to 0 indicate that the aligned signatures are highly likely to originate from different sources.

@@

When evaluating the \citet{aoas2} algorithm, we will describe errors as a percentage of correct evaluations. Statistics that do not translate to percent-form interpretation, such as area under the curve (AUC), will be presented as a decimal.

@@

In this paper, we validate the algorithm described in \citet{aoas2} on three external test sets. We assess the results on each set and evaluate whether the minimal requirements described above are fulfilled for bullet-to-bullet SAM scores. We compare the random forest algorithm's performance to the performance of other suggested measures for quantitative assessment of match strength, such as cross correlation and consecutive matching striae.

@@

The algorithm in \citet{aoas2} is trained on scans from Hamby sets 252 and 173 made available through \citet{nist}. Set 173 was originally published as Hamby set 44. This mis-labeling has been corrected now. The scans used to train the model were taken at 20 fold magnification for a resolution of 1.5625 microns per pixel. While magnification is generally of interest in microscopy, **resolution** -- usually measured in microns per pixel -- is of more interest for 3D topographic measurements, as it determines the operative level of available data.

@@

Case studies were chosen such that difficulty for the matching algorithm increases with each case study: Hamby set 44 is part of the Hamby study. The algorithm in \citet{aoas2} was trained on sets 173 and 252, so the bullets in Hamby 44 are of the same type of ammunition and are fired through the same barrels as the bullets in the training set. The Phoenix PD set uses Ruger P-95 barrels, which are different from the barrels in the Hamby sets used for training the algorithm, but the barrels are rifled similarly to the Ruger P-85 barrels. The Houston sets use Ruger LCP barrels. These barrels are first rifled traditionally, i.e. similar to the Ruger P-85, but are treated to a secondary round of heating after rifling. This second round of heat exposure has the potential to introduce some subclass characteristics, i.e.\ anomalies that are shared between different barrels manufactured in this manner. This should make the automatic matching harder, and in particular, is expected to complicate the classification as different-source land engraved areas/bullets. In all three studies, the scan resolution is much higher than the scans used to train the algorithm in \citet{aoas2}; this difference also provides a test of the algorithm's ability to generalize to scans taken at different resolutions.

@@

At the beginning of this study, we anticipated Hamby 44 would be the easiest set to work with because of its similarity to the sets used to train the random forest algorithm. In a surprise turn of events, it was the hardest to solve, in part because of damage to the bullets that obscured LEA striae. It has been shown \citet{lpr2} that parts of lands can be used for successful identifications. The algorithm proposed by \citet{aoas2} is not capable of  automatically detecting parts of lands with well-expressed striae. It may be useful to couple the \citet{aoas2} algorithm with an algorithm which assesses the quality of the input data and determines which portions of the data to use for comparison. This would emulate the process used by examiners, who first assess the quality of the evidence and whether there is enough information present to attempt a match. One major disadvantage of an automated algorithm is that all of the decisions humans make (recognizing degraded land areas, excluding those areas from consideration, matching only the remaining areas) must be explicitly characterized; however, this explicit characterization means that the process can be scientifically validated to a much higher degree than the human perceptual process.

@@

The random forest matching algorithm presented in \citet{aoas2} does perform well on three different external test sets. While a single cutoff value cannot be used to distinguish matches and non-matches across the different test sets, the algorithm makes no errors when a set-specific cutoff value is used on bullet-to-bullet scores aggregated using sequence average maximum. This performance is on par with the performance of forensic examiners. For a future round of the Houston study, we are planning to compare the algorithmic performance directly with scores given by forensic toolmark examiners. The algorithm's score is not intended to replace an examiner's judgment; instead, it provides a complementary tool to measure and quantify the strength of an identification. 
By validating the algorithm on external test sets, we have demonstrated that the method can be generalized to different types of ammunition and is not overly sensitive to small differences in rifling procedure. Future studies can and should generalize this to a wider range of external test sets to establish the limits of the algorithm's generalizability.
